---
output:
  html_document:
    css: styles.css
---

<br>

## Evaluating Reminder Push Notifications on InstaCart Orders
**Darian Miranda Olivas**

## Overview
End to end A/B testing workflow evaluating whether reminder push notifications increase InstaCart ordering activity using randomized experimental analysis.


## Experiment Findings
Reminder notifications increased orders by approximately **0.198 orders per user** with a **95% confidence interval of 0.193–0.202**, indicating a statistically significant positive treatment effect.

## Data Source
This analysis uses the InstaCart Online Grocery Basket Analysis Data set made available on Kaggle. Access the data [here](https://www.kaggle.com/datasets/yasserh/instacart-online-grocery-basket-analysis-dataset).

<br>
<br>

### Data Integration
**Purpose:** Prepare raw Instacart data into a consistent format ready for analysis.

```{r data_integration, message=FALSE, warning=FALSE}

#====== Set seed for reproducibility =====
  
RNGkind("L'Ecuyer-CMRG")
set.seed(94745850) ##seed number was randomly typed into keyboard





#====== Load in necessary packages ======

library(pacman) #for installing and loading packages

p_load(effectsize, #for Cohens d
       modelsummary, #for model tables
       pwr, #for power test
       rstatix, #for Wilcoxon test
       sandwich, #for robust SEs
       tidyverse) #for code structure





#====== Read in data ======

orders_tbl <- read_csv("orders.csv")

order_products_tbl <- read_csv("order_products__prior.csv")





#====== Join order and product tbls =======

base_tbl <- orders_tbl %>%
  left_join(order_products_tbl, by = "order_id") %>%
  select(-order_hour_of_day, -add_to_cart_order, -reordered) %>%
  filter(eval_set == "prior") #data set originally for prediction so we filter out incomplete train and test evaluation sets
  
```
<br>

### Data Validation
**Purpose:** Ensure data reliability by performing quality checks before constructing the experimental data set.

```{r data_validation, message=FALSE, warning=FALSE}

#====== Inspect data ======

glimpse(base_tbl)





#====== Check for missing values ====

colSums(is.na(base_tbl))

#Result is there exist NA values in the days_since_prior_order column



#---- Inspect missing values ----

is_na_base_tbl <- base_tbl %>%
  filter(is.na(days_since_prior_order) == TRUE)

glimpse(is_na_base_tbl)

#Result is that NA only appears in in days_since_prior_order column when order_number column is 1, this is a structural NA and not missing data





#====== Check number of users and orders =====

n_distinct(base_tbl$user_id)


n_distinct(base_tbl$order_id)





#====== Check statistics for key numeric variables =====

base_tbl %>%
  select(order_number, order_dow, days_since_prior_order) %>%
  summary()





#====== Check for duplicates ======

#anyDuplicated(base_tbl)

#Result is that there are no duplicated rows

```
<br>

### Experiment Design
**Purpose:** Construct user level features and simulate a randomized controlled experiment, including treatment assignment and post treatment outcomes.

```{r experiment_design}

#The summary statistics below would only be calculated for subjects actually in the experiment, however, because I will use the entire data in this simulated experiment and to reduce dimensions for my personal computer, I do this before treatment assignment.


#====== Calculate order sizes ======

order_sizes_tbl <- base_tbl %>%
  select(order_id, product_id) %>%
  group_by(order_id) %>%
  summarise(order_size = n()) %>% 
  ungroup()
  




#====== Calculate user level statistics ======

by_user_tbl <- base_tbl %>%
  select(order_id, user_id, order_dow, days_since_prior_order) %>%
  left_join(order_sizes_tbl,
            by = "order_id") %>%
  group_by(user_id) %>%
  summarise(
    
    prior_total_orders      = n_distinct(order_id),
    avg_days_between_orders = mean(days_since_prior_order, na.rm = TRUE),
    avg_order_size          = mean(order_size),
    weekend_order_rate      = mean(order_dow %in% c(0, 6)) #Sun = 0, Sat = 6
    
    ) %>%
  ungroup()





#====== Randomly assign treatment =====

experiment_tbl <- by_user_tbl %>%
  mutate(group = rbinom(n(), 1, 0.5)) %>%
  mutate(group = factor(group))





#====== Simulate treatment effect ====

experiment_tbl <- experiment_tbl %>%
    mutate(incremental_orders = if_else(group == "1",
                                       rpois(n(), lambda = 0.3),  #treatment lift
                                       rpois(n(), lambda = 0.1)))  #baseline noise





#====== Define frequent users ====

#Frequent users are those in the top 25% of total orders prior to the experiment
#Flagging frequent users will allow us to test for heterogeneous treatment effects later

frequent_user_threshold <- quantile(experiment_tbl$prior_total_orders, 0.75)


experiment_tbl <- experiment_tbl %>%
  mutate(is_frequent_user = if_else(prior_total_orders >= frequent_user_threshold, 1, 0))  

```
<br>

### Estimating Sample Size (n)
**Purpose:** Estimate the required sample size to detect a meaningful treatment effect with sufficient statistical power.

```{r n_estimate}

#In practice this would be estimated before an experiment takes place


#====== Estimate pre experiment Cohens d ====

#I use prior total order variability as a proxy for incremental orders outcome variance

sd_prior_total_orders <- sd(experiment_tbl$prior_total_orders)

delta <- 0.5 #business decision

pre_cohens_d <- delta/sd_prior_total_orders





#====== Estimate required sample size ====

pwr.t.test(d           = pre_cohens_d,
           power       = 0.8,
           sig.level   = 0.05,
           type        = "two.sample",
           alternative = "two.sided")

#Result is that we need at least 17,431 subjects in each group in order to reliably detect a meaningful change in behavior


nrow(experiment_tbl)

```
<br>

### Visualizing Groups
**Purpose:** Compare outcome distributions across treatment and control groups to assess distributional properties and validate modeling assumptions.

```{r visualizing_groups}

#I use histograms instead of density plots for incremental_orders and total_orders because these are discrete count variables.

#====== Create incremental orders histogram =====

plot_incremental_orders <- ggplot(experiment_tbl, aes(x = incremental_orders, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.4) +
  
  #---- specify colors for plots ----
  scale_fill_manual(
    values = c("0" = "#4C78A8",
               "1" = "#F58518"),
    labels = c("Control",
               "Treatment")) +
  
  #---- specify titles and labels ----
  labs(
    title = "Distribution of Incremental Orders by Group",
    x = "Incremental Orders",
      y = "Number of Users",
    fill = NULL
  ) +
  
  #---- specify theme ----
  theme_minimal(base_size = 12) +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "top"
  )

plot_incremental_orders





#====== Create prior total orders histogram =====

plot_prior_total_orders <- ggplot(experiment_tbl, aes(x = prior_total_orders, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.4) +
  
  #---- specify colors for plots ----
  scale_fill_manual(
    values = c("0" = "#4C78A8",
               "1" = "#F58518"),
    labels = c("Control",
               "Treatment")) +
  
  #---- specify titles and labels ----
  labs(
    title = "Distribution Of Prior Total Orders",

    x = "Prior Total Orders",
    y = "Number of Users",
    fill = NULL
  ) +
  
  #---- Specify theme ----
  theme_minimal(base_size = 12) +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "top"
  )

plot_prior_total_orders





#====== Create avg days between orders density graph =====

plot_avg_days_between_orders <- ggplot(experiment_tbl, aes(x = avg_days_between_orders, fill = group)) +
  geom_density(alpha = 0.4, linewidth = 0.2) +
  
  #---- specify colors for plots ----
  scale_fill_manual(
    values = c("0" = "#4C78A8",
               "1" = "#F58518"),
    labels = c("Control",
               "Treatment")) +
  
  #---- specify titles and labels ----
  labs(
    title = "Distribution Of Average Days Between Orders",
    x = "Avg Days Between Orders",
    y = "Density",
    fill = NULL
  ) +
  
  #---- Specify theme ----
  theme_minimal(base_size = 12) +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "top"
  )

plot_avg_days_between_orders





#====== Create avg order size density graph =====

plot_avg_order_size <- ggplot(experiment_tbl, aes(x = avg_order_size, fill = group)) +
  geom_density(alpha = 0.4, linewidth = 0.2) +
  
  #---- specify colors for plots ----
  scale_fill_manual(
    values = c("0" = "#4C78A8",
               "1" = "#F58518"),
    labels = c("Control",
               "Treatment")) +
  
  #---- specify titles and labels ----
  labs(
    title = "Distribution of Average Order Size",
    x = "Avg Order Size",
    y = "Density",
    fill = NULL
  ) +
  
  #---- Specify theme ----
  theme_minimal(base_size = 12) +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "top"
  )

plot_avg_order_size





#====== Create weekend order rate density graph =====

plot_weekend_order_rate <- ggplot(experiment_tbl, aes(x = weekend_order_rate, fill = group)) +
  geom_density(alpha = 0.4, linewidth = 0.2) +
  
  #---- specify colors for plots ----
  scale_fill_manual(
    values = c("0" = "#4C78A8",
               "1" = "#F58518"),
    labels = c("Control",
               "Treatment")) +
  
  #---- specify titles and labels ----
  labs(
    title = "Distribution of Weekend Order Rate",
    x = "Weekend Order Rate",
    y = "Density",
    fill = NULL
  ) +
  
  #---- Specify theme ----
  theme_minimal(base_size = 12) +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "top"
  )

plot_weekend_order_rate


#Overall treatment and control distributions appear broadly similar, which is consistent with successful randomization

```
<br>

### Balance Tests
**Purpose:** Evaluate whether treatment and control groups are statistically similar across pre treatment covariates, verifying validity of random assignment.

```{r balance_tests}

#====== Test for statistical differences across key user level statistics =====

balance_prior_total_orders <- lm(prior_total_orders ~ group, data = experiment_tbl)

balance_avg_days_between_orders <- lm(avg_days_between_orders ~ group, data = experiment_tbl)

balance_avg_order_size <- lm(avg_order_size ~ group, data = experiment_tbl)

balance_weekend_order_rate <- lm(weekend_order_rate ~ group, data = experiment_tbl)





#====== Display model results in a single table ====

modelsummary(
  list(
    "Prior Total Orders"       = balance_prior_total_orders,
    "Avg Days Between Orders"  = balance_avg_days_between_orders,
    "Avg Order Size"           = balance_avg_order_size,
    "Weekend Rate"             = balance_weekend_order_rate
    ),
  stars       = TRUE,
  vcov        = "HC1", #use robust standard errors
  statistic   = "std.error",
  coef_rename = c("group1" = "Difference (Treatment - Control)"),
  gof_omit    = "R2|Adj|AIC|BIC|Log|F|RMSE",
  title       = "Pre Treatment Covariate Balance Assessment"
)

```
<br>

### Experiment Test
**Purpose:** Estimate the causal effect of the treatment using regression and assess robustness of results.
```{r experiment_test}

#====== Test statistical difference across groups for incremental orders using linear model ====

test_model <- lm(incremental_orders ~ group * is_frequent_user +
                 prior_total_orders +
                 avg_days_between_orders +
                 avg_order_size +
                 weekend_order_rate,
                 data = experiment_tbl)





#====== Display results in formatted table =====

modelsummary(test_model,
             stars       = TRUE,
             vcov        = "HC1", #use robust standard errors
             statistic   = "conf.int",
             coef_rename = c("group1"                    = "Treatment",
                             "is_frequent_user"          = "Frequent User",
                             "prior_total_orders"        = "Prior Total Orders",
                             "avg_days_between_orders"   = "Avg Days Between Orders",
                             "avg_order_size"            = "Avg Order Size",
                             "weekend_order_rate"        = "Weekend Order Rate",
                             "group1:is_frequent_user"   = "Frequent User Treatment Effect"),
             gof_omit    = "R2|Adj|AIC|BIC|Log|F|RMSE",
             title       = "Estimated Treatment Effects of Push Notifications On User Ordering Behavior")





#I include the Wilcoxon test as a robustness check, as it does not rely on distributional assumptions and is less sensitive to outliers

#====== Test statistical difference across groups for incremental orders using Wilcoxon test ====

wilcox.test(incremental_orders ~ group,
            data        = experiment_tbl,
            conf.int    = TRUE,
            conf.level  = 0.95,
            exact       = FALSE,
            correct     = FALSE,
            alternative = "two.sided")





#====== Estimate effect size of experiment ====

cohens_d <- cohens_d(incremental_orders ~ group, data = experiment_tbl)





#====== Estimate power of Wilcoxon test results ====

min_group_n <- floor(nrow(experiment_tbl) / 2) #groups are uneven by 1 subject since total participants was an odd number

pwr.t.test(d           = cohens_d$effsize, 
           sig.level   = 0.05,
           n           = min_group_n,
           type        = "two.sample",
           alternative = "two.sided")

```
<br>

### Presentation
**Purpose:** Visualize outcome distributions across treatment and control groups and present test results.

```{r presentation}

#====== Create conclusive incremental orders histogram =====

plot_incremental_orders <- ggplot(experiment_tbl, aes(x = incremental_orders, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.4) +
  
  #---- specify colors for plots ----
  scale_fill_manual(
    values = c("0" = "#4C78A8",
               "1" = "#F58518"),
    labels = c("Control",
               "Treatment")) +
  
  #---- specify y axis scale ----
  scale_y_continuous(labels = function(x) x / 1000) +
  
  #---- specify titles and labels ----
  labs(
    title = "Distribution of Incremental Orders by Group",
    subtitle = "Reminder notifications increase orders by ~0.198 per user (95% CI: 0.193–0.202)",
    x = "Incremental Orders",
    y = "Number of Users (thousands)",
    fill = NULL
  ) +
  
  #---- Specify theme ----
  theme_minimal(base_size = 12) +
  
  theme(
    panel.grid = element_blank(), ##remove grid lines
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "top"
  )

plot_incremental_orders


ggsave("plot_incremental_order.png", plot_incremental_orders, width = 10, height = 6, dpi = 200)

```
